---
title: "mirDriver"
author: "Banabithi Bose, Matt Moravec,  Serdar Bozdag"
date: "5/12/2020"
output: rmarkdown::html_vignette
r_packages: rmarkdown
vignette: > 
    %\VignetteIndexEntry{mirDriver} 
    %\VignetteEngine{knitr} 
    %\VignetteEncoding[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Table of Contents
    * [Introduction](#Introduction)
    * [Example Data Sets](#Example-Data-Sets)
        * [gistic bedfile](#gistic-bedfile)
        * [mirna bedfile](#mirna-bedfile)
        * [RNA seq count data](#RNA-seq-count-data)
        * [ovary cnv masked data](#ovary-cnv-masked-data)
        * [lesions file](#lesions-file)
        * [genes coordinate bedfile](#genes-coordinate-bedfile)
        * [hg38 gene coordinate bedfile](#hg38-genes-coordinate-bedfile)
        * [CNVData](#CNVData)
        * [mirnaData](#mirnaData)
        * [RSeqData](#RSeqData)
        * [TFData](#TFData)
        * [methylationData](#methylationData)
    * [Using the Functions](#Using-the-Functions)
        * [getRegionWiseGistic](#getRegionWiseGistic)
        * [getDifferentiallyExpressedGenes](#getDifferentiallyExpressedGenes)
        * [makingCisAndTransGenes](#makingCisAndTransGenes)
        * [gatherCisGenes](#gatherCisGenes)
        * [getTransGenePredictorFile](#getTransGenePredictorFile)
        * [lassoParallelForTransGene](#lassoParallelForTransGene)
    * [Example Workflow Step 1](#Example-Workflow-Step-1)
    * [Example Workflow Step 2](#Example-Workflow-Step-2)
    * [Example Workflow Step 3](#Example-Workflow-Step-3)
    * [Example Workflow Step 4](#Example-Workflow-Step-4)
    * [Combined Example Workflow](#Combined-Example-Workflow)
    * [Summary](#Summary)
    * [Acknowledgements](#Acknowledgements)

## <a id = "Introduction"></a> Introduction

The miRDriver package is a tool for infering copy number derived miRNA gene networks in cancer. This four step process allows the user to start with copy number variations to find the differentially expressed genes and miRNA that are associated with each trans gene. Then using the differentially expressed genes and miRNA run a lasso regression with factors of methylation, transcription factors, miRNA expression, RNA sequence, and copy number variation data to find which miRNA contribute significantly. The advantage to this tool over other tools that do the same thing is that miRDriver is more accurate compared to them.

# Load miRDriver

Loading the package will make it usable and check to make sure that all of the nessecary packages are also installed. 

```{r}
library("miRDriver");
```

## <a id = "Example-Data-Sets"></a> Example Data Sets
The example datasets are included on a github, https://github.com/bozdaglab/miRDriver-Datasets. All that are small enough are also included in the package. The following sections are to show what these data sets look like. Some included datasets are just examples of the entire datasets and can not be used to run in the code.
## <a id = "gistic-bedfile"></a> gistic bedfile

```{r}
load(system.file("extdata", "gistic_bedfile.rda", package="miRDriver"))
print(region[1:3,1:4])
```

## <a id = "mirna-bedfile"></a> mirna bedfile
```{r}
load(system.file("extdata", "mirna_bedfile.Rda", package="miRDriver"))
print(mirna_bedfile[1:3,1:3])
```

## <a id = "RNA-seq-count-data"></a> RNA seq count data
```{r}
load(system.file("extdata", "RnaSeq_count_dataSmall.Rda", package="miRDriver"))
print(RnaSeq_count_dataSmall)
```
## <a id = "ovary-cnv-masked-data"></a> ovary cnv masked data
```{r}
load(system.file("extdata", "CNVDataSmall.Rda", package="miRDriver"))
print(CNVDataSmall)
```
## <a id = "lesions-file"></a> lesions file
```{r}
load(system.file("extdata", "allLesionsSmall.Rda", package="miRDriver"))
print(allLesionsSmall)
```
## <a id = "genes-coordinate-bedfile"></a> genes coordinate bedfile
```{r}
load(system.file("extdata", "Genes_coord_bedfile.rda", package="miRDriver"))
print(Genes_coord[1:10,1:4])
```
## <a id = "hg38-genes-coordinate-bedfile"></a> hg38 genes coordinate bedfile
```{r}
##load("~/mirDriver/inst/extdata/hg38_genes_coord_bedfile.rda")

load(system.file("extdata", "hg38_genes_coord_bedfile.rda", package="miRDriver"))

print(Genes_coord[1:10,1:4])
```
## <a id = "CNVData"></a> CNVData
```{r}
load(system.file("extdata", "CNVDataSmall.Rda", package="miRDriver"))
print(CNVDataSmall)
```

## <a id = "mirnaData"></a> mirnaData
```{r}
load(system.file("extdata", "mirnaData.Rda", package="miRDriver"))
print(mirnaData[1:10,1:10])
```
## <a id = "RSeqData"></a> RSeqData
```{r}
load(system.file("extdata", "RSeqData.Rda", package="miRDriver"))
print(RSeqDataSmall)
```
## <a id = "TFData"></a> TFData
```{r}
load(system.file("extdata", "TFData.Rda", package="miRDriver"))
print(TFData[1:10,1:2])
```
## <a id = "methylationData"></a> methylationData
```{r}
load(system.file("extdata", "methylationDataSmall.Rda", package="miRDriver"))
print(methylationDataSmall)
```

## <a id = "Using-the-Functions"></a> Using the Functions

* getRegionWiseGistic
* getDifferentiallyExpressedGenes
* makingCisAndTransGenes
* gatherCisGenes
* getTransGenePredictorFile
* lassoParallelForTransGene

The functions make up the functionality of the four steps. Some are used in conjunction to complete a step while others perform the entire step by themselves.



## <a id = "getRegionWiseGistic"></a> getRegionWiseGistic

This function will find the peak wise files based on a gistic file provided by the user. This can be produced on GenePattern.com manually or by other means. The other argument is a logic value for whether or not the user wants to produce a bedfile from this function. To compute the gistic file one can use the gistic2.0 module on Genepattern.com. The link is, https://cloud.genepattern.org/gp/pages/index.jsf?lsid=urn:lsid:broad.mit.edu:cancer.software.genepattern.module.analysis:00125:6.15.28 .


Arguments:
gisticfile - a character string for the memory location of the gistic file or                       amplification
gistic_bedfile - a logical value for whether or not one wants a gistic_bedfile


gisticfile<-"~/mirDriverFold/EXAMPLE_DATASETS/all_lesions.conf_90.txt"
getRegionWiseGistic(gisticfile=gisticfile,gistic_bedfile=TRUE)


## <a id = "getDifferentiallyExpressedGenes"></a> getDifferentiallyExpressedGenes

This function finds the differentially expressed genes based off of peak wise files and RNA sequence count data. This function can be run in parallel or on one core. The first argument one specifies the number of cores one wants to use, the second argument is for the RNA sequence count data, and the third is a logic value for whether or not one wants to run the code in parallel. The peakwise files should be in the correct folder to be used with the code,  "~/mirDriver/miRDriver_Step1/Regoinwise_Gistic_Files". 

Arguments:
ncore - assuming parallel core use, an integer value for the number of cores user wants         to use
RNAcount - a R dataframe of RNA seq count
parallel - a logic value for whether or not one wants to run the code in parallel, if             one is running this code on windows then parallel must be false 

ncore<- 65
load("~/mirDriverFold/EXAMPLE_DATASETS/RnaSeq_count_data.Rda")
getDifferentiallyExpressedGenes(ncore, RNACount=RnaSeq_count_data,parallel = TRUE)


## <a id = "makingTransAndCisGenes"></a> makingTransAndCisGenes


This function makes all of the trans and cis genes for the differentially expressed genes. The input to this function is a gene coordinate bedfile and a gistic bedfile. A bedfile is a dataframe that includes the chromosome, start, end and peak for each gene. 
Arguments:
Genes_coord_bedfile - A dataframe that fits the format of a bedfile for gene                                 coordinates
gistic_bedfile - A dataframe that fits the format of a bedfile for the gistic results                   with amplification/deletion peaks with patients


load("~/mirDriverFold/EXAMPLE_DATASETS/hg38_genes_coord_bedfile.rda")
load("~/mirDriverFold/EXAMPLE_DATASETS/gistic_bedfile.rda")
makingCisAndTransGenes(Genes_coord_bedfile=Genes_coord, gistic_bedfile=region)


## <a id = "gatherCisGenes"></a> gatherCisGenes


This function gathers all of the cis genes for each trans gene. The inputs, gistic bedfile and miRNA, are used to collect of the miRNA and differentially expressed genes.

Arguments:
ncores - Assuming parallel, and integer value that represents the number of cores that          the user wants to use
mirna_bedfile - A dataframe in the format of a bedfile for all of the micro RNA
gistic_bedfile - A dataframe in the format of a bedfile for the gistic results
parallel - A logic value for whether or not one wants to run the code in parallel, if             running code on windows, must be false

load("~/mirDriverFold/EXAMPLE_DATASETS/mirna_bedfile.Rda")
load("~/mirDriverFold/EXAMPLE_DATASETS/gistic_bedfile.rda")
ncores<- 65
gatherCisGenes(ncores,mirna_bedfile=mirna_bedfile,gistic_bedfile=region,parallel=TRUE)


## <a id = "getTransGenePredictorFile"></a> getTransGenePredictorFile


This function creates the trans gene predictor file which is a dataframe with all of the contributors to the expression including methylation data, RNA sequence data, copy number varaition data, transcription factor data and mirna data. This function can be run in parallel or on a single core depending on user preference and machine capabilities.

Arguments:
ncores - assuming parallel use, the number of cores the user wants to use
methylationData - a dataframe for the methylation data
RSeqData - a dataframe for the RNA sequence data
CNVData - a dataframe for the copy number varation data
TFData - a dataframe for the transcription factor data
mirnaData - a dataframe for the mirna data
parallel - a logic value for whether or not the user wants to run the code in parallel,            if running the code on windows, must be false

ncores<- 65
load("~/mirDriverFold/EXAMPLE_DATASETS/methylationData.Rda")
load("~/mirDriverFold/EXAMPLE_DATASETS/RSeqData.Rda")
load("~/mirDriverFold/EXAMPLE_DATASETS/CNVData.Rda")
load("~/mirDriverFold/EXAMPLE_DATASETS/TFData.Rda")
load("~/mirDriverFold/EXAMPLE_DATASETS/mirnaData.Rda")
getTransGenePredictorFile(ncores, methylationData, RSeqData, CNVData, TFData, mirnaData,parallel = TRUE)


## <a id = "lassoParallelForTransGene"></a> lassoParallelForTransGene

This function runs a lasso regression to find what contributes to the expression of the trans gene. This type of regression eliminate non-contributors by settting their coefficient to zero. If the coeffiecient of a miRNA is non-zero after the regression finishes then it is a contributor. However, to be accurate the regression is set to run multiple times (the user specifies as an argument), with a certain percentage of times needed with the coefficient to be non-zero. (the user specifies this as an argument as well). The user also needs to provide the number of folds they want their regression to use.

Arguments:
ncores - assuming the code in run in parallel, an integer for the number of cores one wants to use
Nfolds - an integer for the number of times the lasso regression is run
numCounter - an integer for the number of folds one wants for their lasso regression
nonZeroPercent - an integer for the percentage of appearance as a non-zero LASSO                        coefficient for a mirna to be selected
parallel - a logic value for whether or not the user wants to run the code in parallel,            if the code is run on windows, must be false

ncores=65
Nfolds<-3
numCounter = 20 
nonZeroPercent = 10 
x<-lassoParallelForTransGene(ncores = 70,numCounter = 3,Nfolds = 3,nonZeroPercent = 10,parallel = TRUE)


## <a id = "Example-Workflow-Step-1"></a> Example Workflow Step 1


Step one of the mirDriver pipeline is to find frequently aberatted genes among patients. The package includes multiple functions to perform this step, but the user only needs to one. The user should use getRegionWiseGistic1 if they are starting with copy number variation data and getRegionWiseGistic2 if the user is starting with aberattion data. The code examples below are examples, but do not contain any output. To see output see individual functions in the above sections.

Using getRegionWiseGistic1

humanGene<- "Human_Hg38.UCSC.add_miR.160920.refgene.mat"
copyNumber<- "https://cloud.genepattern.org/gp/users/miRDriver/ovary_brca_cnv_masked.txt"
conf.lev<-0.90
username <- "miRDriver"
password <- "1234"

Using getRegionWiseGistic2

gisticfile<-"~/mirDriver_package_content/EXAMPLE_DATASETS/all_lesions.conf_90.txt"
getRegionWiseGistic2(gisticfile=gisticfile,gistic_bedfile=TRUE)

## <a id = "Example-Workflow-Step-2"></a> Example Workflow Step 2


Step two of the mirDriver pipeline is to find the differentially expressed cis and trans genes. To complete this step one must use two functions, getDifferentiallyExpresssedGenes and mkaingTransAndCisGenes.The code examples below are examples, but do not contain any output. To see output see individual functions in the above sections.


ncore<- 65
load("~/mirDriver_package_content/EXAMPLE_DATASETS/RnaSeq_count_data.Rda")
getDifferentiallyExpressedGenes(ncore, RNACount=RnaSeq_count_data,parallel = TRUE)
load("~/mirDriver_package_content/EXAMPLE_DATASETS/hg38_genes_coord_bedfile.rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/gistic_bedfile.rda")
makingCisAndTransGenes(Genes_coord_bedfile=Genes_coord, gistic_bedfile=region)


## <a id = "Example-Workflow-Step-3"></a> Example Workflow Step 3


Step three of the mirDriver pipeline is to gather the cis genes and cis miRNA for each trans gene. This step is comprised of one function, gatherCisGenes. The code examples below are examples, but do not contain any output. To see output see individual functions in the above sections.


load("~/mirDriver_package_content/EXAMPLE_DATASETS/mirna_bedfile.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/gistic_bedfile.rda")
ncores<- 65
gatherCisGenes(ncores,mirna_bedfile=mirna_bedfile,gistic_bedfile=FALSE,parallel=TRUE)


## <a id = "Example-Workflow-Step-4"></a> Example Workflow Step 4

Step four of the mirDriver pipeline is to run a lasso regression on the data to find which miRNA are indicators for the trans gene expression. This step is comprised of two functions, getTransGenePredictorFile and lassoParallelForTransGene. The code examples below are examples, but do not contain any output. To see output see individual functions in the above sections.


ncores<- 65
load("~/mirDriver_package_content/EXAMPLE_DATASETS/methylationData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/RSeqData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/CNVData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/TFData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/mirnaData.Rda")
getTransGenePredictorFile(ncores, methylationData, RSeqData, CNVData, TFData, mirnaData,parallel = TRUE)
Nfolds<-3
numCounter = 20 
nonZeroPercent = 10 
x<-lassoParallelForTransGene(ncores = 70,numCounter = 3,Nfolds = 3,nonZeroPercent = 10,parallel = TRUE)


## <a id = "Combined-Example-Workflow"></a> Combined Example Workflow

To complete the entire pipeline then one will go through all four steps. This appends all of the steps together to create the final product. For this workflow I will use getRegionWiseGistic2, however, if one uses getRegionWiseGistic1 then it will look similar but the code for running getRegionWiseGistic2 will be the code for running getRegionWiseGistic1. The code examples below are examples, but do not contain any output. To see output see individual functions in the above sections.


gisticfile<-"~/mirDriver_package_content/EXAMPLE_DATASETS/all_lesions.conf_90.txt"
getRegionWiseGistic2(gisticfile=gisticfile,gistic_bedfile=TRUE)
ncore<- 65
load("~/mirDriver_package_content/EXAMPLE_DATASETS/RnaSeq_count_data.Rda")
getDifferentiallyExpressedGenes(ncore, RNACount=RnaSeq_count_data,parallel = TRUE)
load("~/mirDriver_package_content/EXAMPLE_DATASETS/hg38_genes_coord_bedfile.rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/gistic_bedfile.rda")
makingCisAndTransGenes(Genes_coord_bedfile=Genes_coord, gistic_bedfile=region)
load("~/mirDriver_package_content/EXAMPLE_DATASETS/mirna_bedfile.Rda")
gatherCisGenes(ncores,mirna_bedfile=mirna_bedfile,gistic_bedfile=FALSE,parallel=TRUE)
load("~/mirDriver_package_content/EXAMPLE_DATASETS/methylationData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/RSeqData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/CNVData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/TFData.Rda")
load("~/mirDriver_package_content/EXAMPLE_DATASETS/mirnaData.Rda")
getTransGenePredictorFile(ncores, methylationData, RSeqData, CNVData, TFData, mirnaData,parallel = TRUE)
Nfolds<-3
numCounter = 20
nonZeroPercent = 10
x<-lassoParallelForTransGene(ncores = 70,numCounter = 3,Nfolds = 3,nonZeroPercent = 10,parallel = TRUE)


## <a id = "Summary"></a> Summary

This package is designed to Infer Copy Number Derived miRNA-Gene Networks in Cancer. The four step process breaks up the different parts of the process into chunks for easy use and produces accurate results.

## <a id = "Acknowledgements"></a> Acknowledgements

This work was supported by the National Institute OF General Medical Sciences of the National Institutes of Health under Award Number R35GM133657.
